---
title: Statistical Approach for Pleiotropy Informed and Functional Annotation Tree
  Guided Prioritization of GWAS Results with `*multiGPATree*' Package
author:
  - Aastha Khatiwada^[Department of Biostatistics and Bioinformatics, National Jewish Health, Denver, CO], Ayse Selen Yilmaz^[Department of Biomedical Informatics, The Ohio State University, Columbus, OH], Bethany J. Wolf^[Department of Public Health Sciences, Medical University of South Carolina, Charleston, SC], Maciej Pietrzak^[Department of Biomedical Informatics, The Ohio State University, Columbus, OH], Dongjun Chung^[Pelotonia Institute for Immuno-Oncology, The James Comprehensive Cancer Center, The Ohio State University, Columbus, OH]
output:
  html_document:
    df_print: paged
    number_sections: yes
bibliography: references.bib
link_citations: yes
biblio-style: apalike
suppress-bibliography: no
fontsize: 10pt
geometry: margin=2.75cm
r_build_args: "--compact-vignettes=both"
# vignette: "%\\VignetteIndexEntry{multiGPATree-vignette} %\\VignetteEngine{knitr::rmarkdown}
#   %\\VignetteEncoding{UTF-8}\n"
vignette: >
  %\VignetteIndexEntry{[multiGPATree-vignette-html]}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
Vignetterbuilder: knitr, rmarkdown
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  prompt = TRUE,
  comment = ""
)
```

# Overview {.tabset}

This vignette provides an introduction to the *`multiGPATree`* package. R package *`multiGPATree`* implements the *`multiGPATree`* method, a novel statistical approach to prioritizing risk-associated SNPs and the combinations of functional annotations related to one or more trait risk-associated SNPs. The *`multiGPATree`* approach employs a hierarchical model to integrate GWAS summary statistics for multiple traits and functional annotation information within a unified framework by combining an iterative procedure (EM algorithm [@dempster1977maximum]) and a multivariate decision tree algorithm (MVPART [@de2002multivariate]).

The package can be loaded with the command:
\newline

```{r, message=FALSE, warning=FALSE}
# install.packages("devtools")
# library(devtools)
# devtools::install_github("cran/mvpart")
# library(mvpart)
# devtools::install_github("asthakhatiwada/multiGPATree")
library(multiGPATree)
```


This vignette is organized as follows.
Section [2](#datastructure) discusses the data structure required to implement the *`multiGPATree`* method and section [3](#workflow) describes the workflow to implement the *`multiGPATree`* method. Section [3.1](#fitting) discusses how to fit a *`multiGPATree`* model, section [3.2](#pruning) describes how to prune a large *`multiGPATree`* model, section [3.3](#annotation) describes the *`multiGPATree`* model plot and functional annotation tree, and finally, section [3.4](#association) explains command lines for association mapping and identification of combination of functional annotations for one or more traits using *`multiGPATree`*.

Please feel free to contact Dongjun Chung at *chung.911@osc.edu* or Aastha Khatiwada at *khatiwadaa@njhealth.org* for any questions or suggestions regarding the *`multiGPATree`* package.

# Data structure {#datastructure}

In this vignette, we use the simulated data (*simdata*) provided in the package to fit the *`multiGPATree`* model for post-GWAS analysis. The simulated data includes information for three traits/phenotypes (P1, P2 and P3). In the simulated data, the number of SNPs and the number of functional annotations are set to $M=1000$ and $K=10$, respectively. The GWAS association  $p$-values for the SNPs are stored in *simdata\$gwasPval* and the binary functional annotation information for the SNPs are stored in *simdata\$annMat*, respectively. The number of rows in *simdata\$gwasPval* is assumed to be the same as the number of rows in  *simdata\$annMat* where the $i$-th ($i=1,...,M$) row of *gwasPval* and *annMat*  correspond to the same SNPs.

```{r}
data("simdata")
class(simdata)
names(simdata)
dim(simdata$gwasPval)
dim(simdata$annMat)
head(simdata$gwasPval)
head(simdata$annMat)
```


# Workflow {#workflow}
## Fitting the multiGPATree Model {#fitting}

We are now ready to fit a *`multiGPATree`* model using the GWAS $p$-value  and functional annotation data described above. We can fit the *`multiGPATree`* model with the command:

```{r, message=FALSE}
fit.mGPATree <- multiGPATree(gwasPval = simdata$gwasPval,
                            annMat = simdata$annMat,
                            initAlpha = 0.1,
                            cpTry = 0.005)
fit.mGPATree
```



## Prunning the multiGPATree model {#pruning}

The \texttt{prune()} function will prune the *`multiGPATree`* model using any cp value between 0 and 1 as shown below.

```{r, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
fit.mGPATree.pruned <- prune(fit.mGPATree@fit$P1_P2, cp = 0.20)
fit.mGPATree.pruned
```

## Functional annotation tree {#annotation}

The \texttt{plot()} and \texttt{leaf()} functions will plot the *`multiGPATree`* functional annotation tree and provide information about the leaves (terminal nodes) in the tree as shown below.

```{r, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
plot(fit.mGPATree@fit$P1_P2)
```

```{r, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
leaf(fit.mGPATree@fit$P1_P2)
```


## Association mapping {#association}

For the fitted *`multiGPATree`* model, we can make inferences about SNPs using the \texttt{assoc()} function by: $(1)$ prioritizing risk-associated SNPs for one or more traits, and (2) identifying the leaves of the *`multiGPATree`* model in which one or more trait risk-associated SNPs are located. In this case, the \texttt{assoc()} function returns four columns. The first column contains binary values where 1 indicates that the SNP is marginally associated with the first trait (P1) and 0 indicates otherwise. Similarly, the second column contains binary values where 1 indicates that the SNP is marginally associated with the second trait (P2) and 0 indicates otherwise. The third column contains binary values where 1 indicates that the SNP is jointly associated with both traits (P1 and P2) and 0 indicates otherwise.Finally, the fourth column provides information regarding the leaf in which a SNP is located in the *`multiGPATree`* plot. The \texttt{assoc()} function allows both local (\texttt{fdrControl="local"}) and global FDR controls (\texttt{fdrControl="global"}) and users can set the threshold to be between 0 and 1 using the '\texttt{FDR}' argument. 

For the simulated data, *`multiGPATree`* model identified $4$ SNPs to be jointly associated with traits P1 and P2 at the nominal global FDR level of $0.01$, all of which are located in leaf 5 which can be interpreted as the 4 risk SNPs being simultaneously annotated for both annotations $A5$ and $A6$. Similarly, *`multiGPATree`* model identified $38$ SNPs to be marginally associated with trait P1 at the nominal global FDR level of $0.01$. Of the  $38$ SNPs, 27 SNPs are located in leaf 6 and are simultaneously annotated for annotations $A1$ and $A2$ and 11 SNPs are located in leaf 5 and are simultaneously annotated for $A5$ and $A6$.The following lines of code can be used to investigate association mapping and functional annotation tree for *`multiGPATree`* models.

```{r, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
assoc.mGPATree <- multiGPATree::assoc(fit.mGPATree@fit$P1_P2,
                                      FDR = 0.01, 
                                      fdrControl="global")
head(assoc.mGPATree)
table(assoc.mGPATree$P1_P2)
table(assoc.mGPATree$P1_P2, assoc.mGPATree$leaf)
table(assoc.mGPATree$P1)
table(assoc.mGPATree$P1, assoc.mGPATree$leaf)
table(assoc.mGPATree$P2)
table(assoc.mGPATree$P2, assoc.mGPATree$leaf)

table(assoc.mGPATree$P1, assoc.mGPATree$P2)
table(assoc.mGPATree$P1, assoc.mGPATree$P2, assoc.mGPATree$P1_P2)
```

# References
